---
output:
  word_document: default
  html_document: default
---


# R Markdown 

```yaml

title: "HW 1 - Reading in Data"
author: "DUE: Thursday, September 11 at 11:59pm"
output: word_document

```

```{r include=FALSE}
knitr::opts_chunk$set(comment = "", echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

*(Keep this section as-is from the prompt.)*

## Task

*(Keep this section as-is from the prompt.)*

## Criteria for Success

*(Keep this section as-is from the prompt.)*

---

# Reading in Data

> **Tip:** Your CSV’s columns are `State`, `Abbr`, `SY2010_11`, `SY2011_12`, `SY2012_13`. Use `na.rm = TRUE` whenever you compute summary stats.

---

### 1) Help pages

*(5 points)*

```{r}
# The function is help(). Use it with subset like this:
help("subset")
```

---

### 2) Why R Script / R Markdown?

*(5 points)*

* Reproducible: rerun on updated data with zero manual edits.
* Transparent: code + narrative in one file for easy grading and collaboration.
* Portable: easy to share; same code runs on different machines.

---

### 3) First impressions of Lander book

*(5 points)*
*(Write 2–3 sentences of your own—don’t overthink it.)*

---

### 4) Missing value in R

*(5 points)*

```{r}
# R uses NA to represent missing values:
NA
```

---

### 5) Function to get an object’s class; apply to airquality\$Temp

*(5 points)*

```{r}
class(airquality$Temp)
```

---

### 6) Read in the ACGR data

*(5 points)*

```{r}
ACGR <- read.csv("C:/Users/USER/Documents/HW_1/data.csv", stringsAsFactors = FALSE)
head(ACGR, 6)
```

> If your local filename differs (e.g., has “(1)” in it), rename it to match the line above **or** change the string to your exact filename.

---

### 7) How many rows and columns?

*(5 points)*

```{r}
dim(ACGR)  # rows, columns
```

---

### 8) Which values to report (observations vs variables)?

*(5 points)*
The number of **observations** is `r nrow(ACGR)` (rows). The number of **variables** is `r ncol(ACGR)` (columns).

---

### 9) Last 7 rows

*(5 points)*

```{r}
tail(ACGR, 7)
```

---

### 10) Mean of **2010–11** rates; save as new object

*(5 points)*

```{r}
mean_2010_11 <- mean(ACGR$SY2010_11, na.rm = TRUE); mean_2010_11
```

---

### 11) States with **2011–12** below **that year’s** mean

*(5 points)*

> The prompt says “use the object from the previous question,” but it refers to **2011–12**. To be precise, compute the **2011–12** mean and use it below.

```{r}
mean_2011_12 <- mean(ACGR$SY2011_12, na.rm = TRUE)
subset(ACGR, SY2011_12 < mean_2011_12, select = State)$State
```

---

### 12) Highest **2011–12** rate; store in `grad_max`

*(5 points)*

```{r}
grad_max <- max(ACGR$SY2011_12, na.rm = TRUE); grad_max
```

---

### 13) State with the highest **2011–12** rate (name only)

*(5 points)*

```{r}
subset(ACGR, SY2011_12 == grad_max, select = State)$State
```

---

### 14) Lowest **2011–12** rate; store in `grad_min`

*(5 points)*

```{r}
grad_min <- min(ACGR$SY2011_12, na.rm = TRUE); grad_min
```

---

### 15) State with the lowest **2011–12** rate (name only)

*(5 points)*

```{r}
subset(ACGR, SY2011_12 == grad_min, select = State)$State
```

---

### 16) Largest **change** from 2010–11 to 2012–13; show signed difference

*(5 points)*

```{r}
chg <- transform(ACGR, diff = SY2012_13 - SY2010_11)
chg[which.max(abs(chg$diff)), c("State", "diff")]
```

---

### 17) Median change from 2010–11 to 2012–13

*(5 points)*

```{r}
median(chg$diff, na.rm = TRUE)
```

---

### 18) Did the **median** rate increase or decrease (2010–11 vs 2012–13)?

*(5 points)*

```{r}
med_10 <- median(ACGR$SY2010_11, na.rm = TRUE); med_13 <- median(ACGR$SY2012_13, na.rm = TRUE)
c(median_2010_11 = med_10, median_2012_13 = med_13, change = med_13 - med_10)
```

*(Positive `change` = increase; negative = decrease.)*

---

### 19) Top five **increases** (2010–11 → 2012–13)

*(5 points)*

```{r}
inc <- subset(chg, diff > 0, select = c(State, diff))
head(inc[order(-inc$diff), ], 5)
```

---

### 20) Largest signed change (duplicate of Q16)

*(5 points)*

```{r}
chg[which.max(abs(chg$diff)), c("State", "diff")]
```

---

